---
title: "LinearRegressionMarkdown"
output:
  pdf_document: default
  html_document: default
date: "2024-11-25"
---

Loads needed librarys
```{r, include=FALSE}
library(tidyverse)
library(plotly)
library(htmlwidgets)
library(car)
```

Defines the vars to generates 3d models with different independent variables
```{r, include=FALSE}
independent_vars <- c("AvgGDP", "AvgLifeExpectancy")  # Adjust these variables as needed

dependent_var <- "AvgHappiness_Score"
```

Sources the data from the TablesJoinedOnCountry.R
```{r, include=FALSE}
source("/cloud/project/Scripts/TablesJoinedOnCountry.R")
```


Extracts the dependednt and independent variables
```{r, include=FALSE}
merged_data <- cleaned_data %>%
  select(all_of(c(dependent_var, independent_vars))) %>%
  drop_na()
```


Dynamically builds formula for different variables
```{r, include=FALSE}
formula <- as.formula(paste(dependent_var, "~", paste(independent_vars, collapse = " * ")))

regression_model <- lm(formula, data = merged_data)
```

Adds the predicated data and residuals to the merged data data frame
```{r, include=FALSE}
merged_data <- merged_data %>%
  mutate(
    Predicted = predict(regression_model, .), 
    Residual = !!sym(dependent_var) - Predicted  
  )
```


Defines the accetable range
```{r, include=FALSE}
acceptable_range <- 0.5  
```


Mutates the plot_data frame to include two new columns
1. Difference- Stores the value between the absolute value between the predicated and Actual happiness score
2. Use a case state to the categories an entry into either "Correct", "Near Miss",or "Misclassified" by looking at the value of the difference variable 
```{r, include=FALSE}
merged_data <- merged_data %>%
  mutate(
    Difference = abs(Predicted - !!sym(dependent_var)),  # Calculate absolute difference
    Category = case_when(
      Difference < 0.01 ~ "Perfect",
      Difference <= acceptable_range ~ "Near Miss",
      TRUE ~ "Misclassified"
    ),
    Category = factor(Category, levels = c("Perfect", "Near Miss", "Misclassified"))
  )
```

Sums each category
```{r, include=FALSE}
category_counts <- merged_data %>%
  group_by(Category) %>%
  summarize(Count = n()) %>%
  ungroup()
```


Calculates the fraction (Perfect + Near Miss)/ Misclassified
```{r, include=FALSE}
PerfectNearMiss <- sum(category_counts$Count[category_counts$Category %in% c("Perfect", "Near Miss")])
Misclassified <- category_counts$Count[category_counts$Category == "Misclassified"]

correctPercentage <- PerfectNearMiss / (PerfectNearMiss + Misclassified)

correctPercentage <-  correctPercentage * 100
```

Caculates Performance metrics
```{r, include=FALSE}
summary_model <- summary(regression_model)
rmse <- sqrt(mean(merged_data$Residual^2, na.rm = TRUE))
r_squared <- summary_model$r.squared
```

Creates text to pasted onto the graph
```{r, include=FALSE}
count_text <- paste(
  "Correct", category_counts$Count[category_counts$Category == "Perfect"],
  "\nNear Miss:", category_counts$Count[category_counts$Category == "Near Miss"],
  "\nMisclassified:", category_counts$Count[category_counts$Category == "Misclassified"],
  "\nCorrect Percentage:", round(correctPercentage, 2), "%",
  "\nRMSE:", round(rmse, 2),
  "\nR-squared:", round(r_squared, 2)
)
```

1. Creates a 2d grid with every combinations of points for a 100 x 100 grid or of 10000 different points to model the plane.(note: 100 is not the range but the amount of discrete points the grid will use when plotting a plane)
2. Predicts the happiness score for each point in the grid and adds to a table
```{r, include=FALSE}
# Create a grid for the regression plane
grid <- expand.grid(
  AvgGDP = seq(min(merged_data$AvgGDP, na.rm = TRUE), max(merged_data$AvgGDP, na.rm = TRUE), length.out = 100),
  AvgLifeExpectancy = seq(min(merged_data$AvgLifeExpectancy, na.rm = TRUE), max(merged_data$AvgLifeExpectancy, na.rm = TRUE), length.out = 100)
)
grid$Predicted <- predict(regression_model, newdata = grid)
```


Dynamically sets the plot title and file name
```{r, include=FALSE}
plot_title <- paste("Regression Plane With Interaction: ", paste(independent_vars, collapse = "-"))
```

Generates an interactive 3d graph using plotly
Key points:
1. add_markers- add the data of the actual countries to the graph
2. add_trace- add the plane of best fit the graph
3. add_trace(2)- the next two add_traces add the planes showing wether or not the points are within the accetable range
4. layout- defins the x, y, and z axis labels
5. title- sets title
6. annotations- adds important text to the plot
```{r, include=FALSE}
plot <- plot_ly() %>%
  add_markers(
    data = merged_data,
    x = ~AvgGDP,
    y = ~AvgLifeExpectancy,
    z = ~AvgHappiness_Score,
    color = ~Category,
    colors = c("green", "blue", "red"),  # Matching colors to categories
    marker = list(size = 5, opacity = 0.8),
    name = "Data Points",
    size = 1
  ) %>%
  add_trace(
    data = grid,
    x = ~AvgGDP,
    y = ~AvgLifeExpectancy,
    z = ~Predicted,
    type = "mesh3d",
    opacity = 0.5,
    color = I("lightgray"),  
    showscale = FALSE,
    name = "Regression Plane"
  ) %>% add_trace(
    x = ~grid$AvgGDP, 
    y = ~grid$AvgLifeExpectancy,
    z = ~grid$Predicted + 0.5,
    type = "mesh3d",
    opacity = 0.3,
    color = I("red"),  
    showscale = FALSE,  
    name = "Upper Boundary"
  ) %>%
  add_trace(
    x = ~grid$AvgGDP, 
    y = ~grid$AvgLifeExpectancy,
    z = ~grid$Predicted - 0.5, 
    type = "mesh3d",
    opacity = 0.3,
    color = I("purple"), 
    showscale = FALSE,
    name = "Lower Boundary"
  )%>%
  layout(
    scene = list(
      xaxis = list(title = independent_vars[1]),
      yaxis = list(title = independent_vars[2]),
      zaxis = list(title = dependent_var)
    ),
    title = plot_title,
    annotations = list(
      list(
        x = 0.95,
        y = 0.95,
        xref = "paper",
        yref = "paper",
        text = count_text,
        showarrow = FALSE,
        font = list(size = 10, color = "black"),
        xanchor = "right",
        yanchor = "top"
      )
    )
  )
```

Prints the plot
```{r, include=TRUE}
# Display plot
print(plot)
```

Actual vs Predicted
```{r, include=FALSE}
# Prepare the data for plotting
plot_predictions <- merged_data %>%
  rename(Actual = !!sym(dependent_var)) %>%
  select(Actual, Predicted, Category)

# Create the ggplot
q <- ggplot(plot_predictions, aes(x = Actual, y = Predicted)) +
  geom_point(aes(color = Category, shape = Category), alpha = 0.7, size = 3) +
  # Add the perfect prediction line
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +
  # Add lines for the acceptable range
  geom_abline(intercept = acceptable_range, slope = 1, color = "blue", linetype = "dotted") +
  geom_abline(intercept = -acceptable_range, slope = 1, color = "blue", linetype = "dotted") +
  theme_minimal() +
  labs(
    title = paste("Actual vs. Predicted", dependent_var),
    subtitle = paste("Linear Regression of", paste(independent_vars, collapse = " + "), "on", dependent_var),
    x = paste("Actual", dependent_var),
    y = paste("Predicted", dependent_var),
    color = "Prediction Category",
    shape = "Prediction Category"
  ) +
  scale_color_manual(
    values = c("Perfect" = "green", "Near Miss" = "purple", "Misclassified" = "red"),
    drop = FALSE
  ) +
  scale_shape_manual(
    values = c("Perfect" = 16, "Near Miss" = 17, "Misclassified" = 4),
    drop = FALSE
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  ) +
  guides(
    color = guide_legend(override.aes = list(shape = c(16, 17, 4))),
    shape = guide_legend(override.aes = list(color = c("green", "purple", "red")))
  ) +
  # Add annotations for performance metrics
  annotate(
    "text",
    x = 6,
    y = 4.5,
    label = count_text,
    size = 3,
    color = "black",
    hjust = 0
  )


```

```{r, include=TRUE}
print(q)
```

Saves the files
```{r, include=FALSE}
# Generate HTML and Summary file names based on independent variables
html_file <- paste0("Regression_With_Interaction_", paste(independent_vars, collapse = "_"), ".html")
summary_file <- paste0("Summary_With_Interaction_", paste(independent_vars, collapse = "_"), ".txt")
output_filename <- "3dWithInteractionActual_vs_Predicted_Plot.png"

# Define paths for saving the plot and model summary
plot_path <- "/cloud/project/Models/3dLinearRegression/DataAndGraphs/WithInteraction/"
model_summary_path <- "/cloud/project/Models/3dLinearRegression/DataAndGraphs/WithInteraction/"

# Save the plot as HTML
# Ensure that 'plot' is an interactive plot object (e.g., created with plotly or ggplotly)
saveWidget(plot, file.path(plot_path, html_file), selfcontained = TRUE)

ggsave(
  filename = file.path(plot_path, output_filename),  # Correctly specify the file path
  plot = q,                                          # The ggplot object to save
  width = 10,                                        # Width of the plot in inches (adjust as needed)
  height = 8,                                        # Height of the plot in inches (adjust as needed)
  dpi = 300                                          # Resolution of the plot
)

# Save the model summary to a text file using sink
sink(file.path(model_summary_path, summary_file))
cat("Regression Model Summary with Interaction:\n\n")
print(summary(regression_model))  # Prints the summary of the linear regression model
cat("\nRMSE:", round(rmse, 2))
cat("\nR-squared:", round(r_squared, 2))
sink()  # Close the sink connection
```
