---
title: "LinearRegressionMarkdown"
output:
  pdf_document: default
  html_document: default
date: "2024-11-25"
---


Load necessary libraries
```{r, include=FALSE}
library(tidyverse)
library(caret)
```

Runs the TablesJoinedOnCountry Script to access the data
```{r, include=FALSE}
source("/cloud/project/Scripts/TablesJoinedOnCountry.R")
```

Extracts the data from the cleaned_data variable in the TablesJoinedOnCountry.R script
```{r, include=FALSE}
data <- cleaned_data %>%
  select(AvgHappiness_Score, AvgGDP) %>%
  drop_na()
```


Sets the seed to make sure all the results can be replicated
Splits the data into train and test sets 
```{r, include=FALSE}
set.seed(123)
train_index <- createDataPartition(data$AvgHappiness_Score, p = 0.8, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]
```

Builds the linear regression model using the train_data
```{r, include=FALSE}
model <- lm(AvgHappiness_Score ~ AvgGDP, data = train_data)
```

```{r, include=FALSE}
# Make predictions on the test set
predictions <- predict(model, test_data)
```

Make predictions using the test data and then calculates performance metrics for the model
```{r, include=FALSE}
residuals <- test_data$AvgHappiness_Score - predictions
rmse <- sqrt(mean(residuals^2))

sst <- sum((test_data$AvgHappiness_Score - mean(test_data$AvgHappiness_Score))^2)
sse <- sum(residuals^2)
r_squared <- 1 - (sse / sst)

```

Sets up data frame for plotting the actual vs predicted scores
```{r, include=FALSE}
plot_data <- data.frame(
  Actual = test_data$AvgHappiness_Score,
  Predicted = predictions
)
```

Acceptable range in which we will say the guess was accurate, any prediction from the model outside of this range will be classified as wrongly predicted 
```{r, include=FALSE}
# Define acceptable range
acceptable_range <- 0.5 
```


Mutates the plot_data frame to include two new columns
1. Difference- Stores the value between the absolute value between the predicated and Actual happiness score
2. Use a case state to the categories an entry into either "Correct", "Near Miss",or "Misclassified" by looking at the value of the difference variable 
```{r, include=FALSE}
plot_predictions <- plot_data %>%
  mutate(
    Difference = abs(Predicted - Actual),
    Category = case_when(
      Difference <= 0.1 ~ "Correct",          
      Difference <= acceptable_range ~ "Near Miss",
      TRUE ~ "Misclassified"
    ),
    Category = factor(Category, levels = c("Correct", "Near Miss", "Misclassified"))
  )
```


Sums each category
```{r, include=FALSE}
category_counts <- plot_predictions %>%
  group_by(Category) %>%
  summarize(Count = n()) %>%
  ungroup()
```


Caculates the fraction (Correct + Near Miss)/ Misclassified
```{r, include=FALSE}
correctPercentage <- category_counts[1,2] / ( category_counts[1,2] + category_counts[2,2]) 
correctPercentage <- round(correctPercentage,2)
correctPercentage <- sprintf("%.2f%%", correctPercentage * 100)
```



Creates reusable text to put on the graphs to display information about the model
```{r, include=FALSE}
count_text <- paste(
  "Near Miss:", category_counts$Count[category_counts$Category == "Near Miss"],
  "\nMisclassified:", category_counts$Count[category_counts$Category == "Misclassified"],
  "\nCorrect Percentage:", correctPercentage,
  "\nRMSE:", round(rmse, 2),
  "\nR-squared:", round(r_squared, 2)
)
```


use ggplot to create a graph
Key details:
1. geom_point-plots the actual scores vs the predicted scores as a scatter plot
2. geom_abline- adds 3 lines: one in the line y= x, any points directly on this curve were correctly predicted. The two other lines represent the margin of error, and any point between these two lines is classified as a near miss.
3. Labs- adds label to the grpah
4. scale_color_manual- changes the color of the points based on accuracy of the predictions
5. scale_shape_manual- changes the shape of the points based on accuracy of the predictions
6. Theme- adjust positions, size, and format of some text
7. guides- correctly formats the legend
8. annonate- adds importnat information as text to the graph 
```{r, include=FALSE}
# Create the Actual vs. Predicted scatter plot
p <- ggplot(plot_predictions, aes(x = Actual, y = Predicted)) +
  geom_point(aes(color = Category, shape = Category), alpha = 0.7, size = 3) +
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +  # Perfect prediction line
  geom_abline(intercept = acceptable_range, slope = 1, color = "blue", linetype = "dotted") +
  geom_abline(intercept = -acceptable_range, slope = 1, color = "blue", linetype = "dotted") +
  theme_minimal() +
  labs(
    title = "Actual vs. Predicted AvgHappiness_Score",
    subtitle = "Linear Regression of AvgGDP on AvgHappiness_Score",
    x = "Actual AvgHappiness_Score",
    y = "Predicted AvgHappiness_Score",
    color = "Prediction Category",
    shape = "Prediction Category"
  ) +
  scale_color_manual(
    values = c("Correct" = "green", "Near Miss" = "purple", "Misclassified" = "red"),
    drop = FALSE  
  ) +
  scale_shape_manual(
    values = c("Correct" = 16, "Near Miss" = 17, "Misclassified" = 4),
    drop = FALSE  # Prevent dropping unused levels
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  ) +
  guides(
    color = guide_legend(override.aes = list(shape = c(16, 17, 4))),
    shape = guide_legend(override.aes = list(color = c("green", "purple", "red")))
  ) +
  annotate(
    "text",
    x = 6,  
    y = 4.25,
    label = count_text,
    size = 3,
    color = "black"
  )
```

Prints the graph
```{r, include=True}
print(p)

```



Gets the data points of the train_data for graph vs the line of best of fit, and generates points for upper and lower bounds for a correct predictions
```{r, include=FALSE}
train_predictions <- predict(model, train_data)

train_data <- train_data %>%
  mutate(
    Predicted = train_predictions,
    Upper = Predicted + acceptable_range,
    Lower = Predicted - acceptable_range
  )
```

Finds the correct amount of points within the accetable range
```{r, include=FALSE}
within_range_count <- sum(abs(train_data$Predicted - train_data$AvgHappiness_Score) <= acceptable_range)
within_range_percentage <- (within_range_count / nrow(train_data)) * 100
within_range_percentage_formatted <- sprintf("%.2f%%", round(within_range_percentage, 2))

```

Modify count text
```{r, include=FALSE}
count_text <- paste(
  "\nRMSE:", round(rmse, 2),
  "\nR-squared:", round(r_squared, 2),
  "\nWithin Acceptable Range:", within_range_count,
  "\n(", within_range_percentage_formatted, ")"
)
```


use ggplot to create a graph
Key details:
1. geom_point-plots the the average GDP of a country vs its actual happiness score
2. geom_smooth- plots the line of best fit for the data
3. geom_line- plots the lines for the upper and lower bounds of a nearly correct guess
4. labs- adds labels to graph
5. theme- adjust the formtating of ceritan labels
6. annotate- adds text to the plot 
7. x/ylim- adjust the range of the x and y axis 
```{r, include=FALSE}
g <- ggplot(train_data, aes(x = AvgGDP, y = AvgHappiness_Score)) +
  geom_point(alpha = 0.7, color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  geom_line(aes(y = Upper), color = "blue", linetype = "dotted") +  # Upper acceptable range
  geom_line(aes(y = Lower), color = "blue", linetype = "dotted") +  # Lower acceptable range
  labs(
    title = "Linear Regression",
    subtitle = "Line of Best Fit",
    x = "Average GDP",
    y = "Average Happiness Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  ) +
  annotate(
    "text",
    x = 1.5,  # Position at the far right
    y = 3.5,  # Position at the top
    label = count_text,
    size = 3,
    color = "black",
  ) +
  xlim(0, 2) +
  ylim(0,8.5)
```

Prints the graph
```{r, include=True}
print(g)
```

Saves the graph
```{r, include=FALSE}
# Save the plots
ggsave(
  filename = "/cloud/project/Models/LinearRegression/DataAndGraphs/Actual_vs_PredictedLinearRegression.png",
  plot = p,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  filename = "/cloud/project/Models/LinearRegression/DataAndGraphs/LinearRegressionModel.png",
  plot = g,
  width = 8,
  height = 6,
  dpi = 300
)

# Save the model summary to a text file
sink("/cloud/project/Models/LinearRegression/DataAndGraphs/LinearRegressionOfGDP.txt")
cat("Linear Regression of AvgHappiness_Score on AvgGDP\n\n")
print(summary(model))
sink()
```